---
title: "Pollutant Data Summary"
author: "Faaez Razeen"
date: "2/26/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Pollutant Data

The zip file contains 332 comma-separated-value (CSV) files containing pollution monitoring data for fine particulate matter (PM) air pollution at 332 locations in the United States. Each file contains data from a single monitor and the ID number for each monitor is contained in the file name. For example, data for monitor 200 is contained in the file "200.csv". Each file contains three variables:

* Date: the date of the observation in YYYY-MM-DD format (year-month-day)
* sulfate: the level of sulfate PM in the air on that date (measured in micrograms per cubic meter)
* nitrate: the level of nitrate PM in the air on that date (measured in micrograms per cubic meter)


```{r Loading files into dataframe}
files <- list.files('specdata', full.names = TRUE)
df <- data.frame()
for(file in files){df <- rbind(df, read.csv(file))}
summary(df)
```

### Date Range

```{r Date Calculations}
start_date <- as.Date(df[1, ][['Date']])
end_date <- as.Date(df[dim(df)[1], ][['Date']])
days_difference = end_date - start_date
paste("Values ranging from", start_date, "to", end_date, "with a time difference of ", days_difference, "days which is equal to", days_difference/365, "years.")
```

### Missing Values

Has a lot of missing values. 

```{r Calculating missing values}
missing_percent <- (1 - ((dim(df[complete.cases(df), ])[1])/(dim(df)[1]))) * 100
paste("Percentage of missing data:", round(missing_percent, digits = 2))
```